Classless Java: Tuning Java Interfaces

### 2015-11-25 paper/sections/CaseStudy.tex
Game example

\section{Case Studies}
In this section we conduct three case studies which reveals various advantages
using \mixin annotation. 1st case study provides a easy and straightforward way
to solve the Expression Problem. 2nd case study models a simple game and the
comparison between our approach and others shows that code amount is reduced
tremendously using \mixin. 3rd case study models an embedded DSL for SQL
languages with fluent interfaces and shows that \mixin not only removes verbol
code for defining fluent interfaces but also does implicit refinement along
extension.

\subsection{A Trivial Solution to the Expression Problem}

The \emph{Expression Problem} (EP)~\cite{wadler98expression-problem} is coined
by Wadler about modular extensibility issues in software evolution and has been
a hot topic in programming languages since. The requirements of a solution to
the EP can be summarized as: 
\begin{itemize}
\item \emph{Extensibility in both dimensions}: A solution must allow
  the addition of new data variants and new operations and support
  extending existing operations.
\item \emph{Strong static type safety}: A solution must prevent
  applying an operation to a data variant which it cannot handle using
  static checks.
\item \emph{No modification or duplication}: Existing code must not be
  modified nor duplicated.
\item \emph{Separate compilation and type-checking}: Safety checks or
  compilation steps must not be deferred until link or runtime.
\item \emph{Independent extensibility}~\cite{zenger05independentlyextensible}:
  It should be possible to combine independently developed extensions so that
  they can be used jointly.
\end{itemize}

Today we know of various solutions to the EP that either rely on \emph{new
  programming language features}
% ~\cite{chambers95multimethods, clifton00multiJava, madsen89virtual, nystrom06j,
%   bruce98astatically, McDirmid01jiazzi, garrigue98polymorphic,
%   zenger01extensible, loh06open, wehr11javaGI}, 
or can be used as \emph{design patterns}%~\cite{gof}
in existing languages.
% ~\cite{torgersen04theexpression, oliveira06extensible,
%  wouter08datatypes, oliveira09modular, oliveira12extensibility}. 
Now we show that the EP can be solved with our \mixin annotation in a trivial
way. We will use the canonical example of implementing arithmetic expressions
gradually to illustrate the usage.

\paragraph{Initial System}
The initial system expresses arithmetic expressions with only literals and
addition structure:

@@

\texttt{Exp} is the common super-interface with an evaluation operation
\texttt{eval()} inside. Sub-interfaces \texttt{Lit} and \texttt{Add} extend
interface \texttt{Exp} with default implementations for \texttt{eval}
operation. The number field (\texttt{x}) of a literal is represented as a getter
method \texttt{x()} and expression fields (\texttt{e1,e2}) of an addition as
getter methods \texttt{e1()} and \texttt{e2()}.

\paragraph{Adding a New Variant}
It is easy to add new data variants to the code in the initial system in a
modular way. For example, the following code shows how to add the subtraction
variant.

@@

\paragraph{Adding a New Operation}
Adding new operations to the previous system is still straightforward (although
requires longer code). The following code shows an example of adding a new
operation \texttt{print}.

\lstinputlisting[linerange=57-66]{../UseMixinLombok/src/test/TestExpression.java}% APPLY:linerange=EXPRESSION_PRINT

The basic idea is to define interfaces for extending old interfaces.  The
interface \texttt{ExpP} extending interface \texttt{Exp} is defined with the
extra function \texttt{print()}. Interfaces \texttt{LitP} and \texttt{AddP} are
defined with default implementations of operation \texttt{print()}, extending
base interfaces \texttt{Lit} and \texttt{Add}, respectively. Importantly, note
that in \texttt{AddP}, the types of ``\emph{fields}'' \texttt{e1,e2} are refined
via the \emph{return types refinement} of getter methods \texttt{e1()} and
\texttt{e2()}.

\paragraph{Independent Extensibility}
To show that our approach supports independent extensibility, we first define a
new operation \texttt{collectLit}(which collects all
literal components in an expression) on expressions:

@@

\subsection{A Small Game Example}
We now present a variant of a classic game example, which is often used to
evaluate reuse ability related to inheritance and design patterns. The code
using our approach and the solution by Bono et. al~\cite{bono14} in online, for
space constraints, we will omit it here. In the game, there is a player with the
goal of collecting as many coins as possible. She may enter a room with several
doors to be chosen among. This is a good example because it involves code reuse
(different kinds of doors inherit a common type, with different features and
behaviour), multiple inheritance (a special kind of door may require features
from two other door types). It also shows how to model symmetric sum, override
and alias as trait-oriented programming. Moreover, the code is provided in
~\cite{bono14} so that we can have a clear and fair comparison with our
approach, as shown in Table~\ref{table:codereduction}.

\begin{table}[h]
\centering
\begin{tabular}{ccc}
\hline
            & SLOC   & \# of classes/interfaces \\ \hline
Bono et al. & 335    & 14                       \\
Ours        & 199    & 11                       \\
\rowcolor[HTML]{C0C0C0} 
Reduced by  & 40.6\% & 21.4\%                   \\ \hline
\end{tabular}
\caption{Code Reduction and Reuse.}
\label{table:codereduction}
\end{table}

\subsection{Embedded DSLs with Fluent Interfaces}
Since the style of fluent interfaces was invented in Smalltalk as method
cascading, more and more languages came to support fluent interfaces, including
JavaScript, Java, C++, D, Ruby, Scala, etc. For most languages, to create fluent
interfaces, programmers have to either handwritten everything or create a
wrapper around the original non-fluent interfaces, using \textbf{this}. In Java,
several libraries help create fluent APIs easily (by code generation), including
jOOQ, op4j, fluflu, JaQue, etc. These libraries have given us hints on the
implementation of fluent setters in ClassLess Java.

When creating fluent interfaces, our approach has two main advantages:
\begin{enumerate}
\item Instead of forcing programmers to hand write code using \textbf{return
    this}, our approach with \mixin annotation removes this verbosity and
  automatically generate fluent setters.
\item Along the extension direction, the return types of fluent setters are
  automatically refined.
\end{enumerate}

\noindent We use embedded DSLs of two simple SQL query languages to illustrate.
The first query language \texttt{Database} (conceptually) models
\texttt{select}, \texttt{from} and \texttt{where} clause:
\lstinputlisting[linerange=6-13]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_DATABASE

\noindent Note that we are only illustrating the idea of modeling embedded DSLs
with the database example, no concrete implementation of database connection nor
operations are provided in this case study. One benefit that fluent methods give
us is method chaining convenience. So when making a query using \texttt{select},
\texttt{from} and \texttt{where}, instead of writing code like:
\lstinputlisting[linerange=26-29]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=NORMAL_QUERY

\noindent We can write more concise code below:
\lstinputlisting[linerange=35-38]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_QUERY1

\noindent Now we extend the first query language with a new feature
\texttt{orderBy} which orders the result records by a field that users
specified. The key is to extend interface \texttt{Database} with only the
feature we need to add and the return type of fluent setters in
\texttt{Database} is automatically refined to \texttt{ExtendedDatabase}:
\lstinputlisting[linerange=17-20]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_DATABASE_EXT

% \lstinputlisting[linerange=44-48]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_QUERY2

>>>>

\section{Case Studies}
In this section we conduct three case studies which reveals various
advantages using \mixin annotation. The first case study provides a
simple way to solve the Expression Problem, while supporting multiple,
independent extensions in Java.  The second case study show how to model an
embedded DSL for SQL languages with fluent interfaces. Finally, the third 
case study models a simple game, and compares our implementation with an 
existing one, showing that the amount of code is reduced significantly using \mixin.

\subsection{A Trivial Solution to the Expression Problem with Object Interfaces}

The \emph{Expression Problem} (EP)~\cite{wadler98expression-problem}
is a well-known problem about modular extensibility issues in software
evolution. Recently, a new solution~\cite{eptrivially} that uses only
covariant type refinement that has been shown to work in Java. When
that solution is modelled with interfaces and default methods, it can
even provide independent extensibility: that is, the ability to
assemble a system, from multiple, independently developed extensions.
Unfortunatelly, a plain Java solution has alot of code verbosity in
Java due to code required for instantiation, which makes the solution 
cumbersome to use. The \mixin annotation helps to remove such code, and 
makes the solution quite easy to use.
\bruno{We should try to do some space saving in here. You could, for example 
combine the code for the initial system with the code for a new operation in a figure/table 
with 2 colums. Then just refer to the Table/Figure in the text.}

\paragraph{Initial System}
In the formulation of the EP there is an initial system that models
arithmetic expressions with only literals and addition, and an initial 
operation:

@@

\noindent \texttt{Exp} is the common super-interface with an evaluation operation
\texttt{eval()} inside. Sub-interfaces \texttt{Lit} and \texttt{Add} extend
interface \texttt{Exp} with default implementations for the \texttt{eval}
operation. The number field \texttt{x} of a literal is represented as a getter
method \texttt{x()} and expression fields (\texttt{e1} and \texttt{e2}) of an addition as
getter methods \texttt{e1()} and \texttt{e2()}.

\paragraph{Adding a New Type of Expressions}
It is easy to add new types of expressions to the code in the initial system in a
modular way. For example, the following code shows how to add subtraction.

@@

\paragraph{Adding a New Operation} The difficulty of the EP in object-oriented 
languages arizes from adding new operations. For example, adding a pretty printing 
operation, would typically involve changing all of the existing code. However, a solution 
to the EP forbids this and it also forbids using casts. In other words, it should be possible 
to add the pretty printing operation in a type-safe and modular way.  
It turns out that, this can be done easily, using our object interfaces approach. 
The following code shows how to add the new operation \texttt{print}.

\lstinputlisting[linerange=57-66]{../UseMixinLombok/src/test/TestExpression.java}% APPLY:linerange=EXPRESSION_PRINT
\bruno{Please add a Java comment to the code emphasizing that the types of the getters are refined!}
\bruno{please indent code properly (the getters are not aligned with default))}

\noindent  The interface \texttt{ExpP} extending interface \texttt{Exp} is defined with the
extra method \texttt{print()}. Interfaces \texttt{LitP} and \texttt{AddP} are
defined with default implementations of \texttt{print()}, extending
base interfaces \texttt{Lit} and \texttt{Add}, respectively. Importantly, note
that in \texttt{AddP}, the types of the ``\emph{fields}'' (i.e. the getter methods) \texttt{e1} 
and \texttt{e2} are refined. If the types were not refined then the body of the \texttt{print}
method in \texttt{AddP} would fail to type-check.

\paragraph{Independent Extensibility}
To show that our approach supports independent extensibility~\cite{odersky}, we first define a
new operation \texttt{collectLit} (which collects all
literal components in an expression) on expressions. For space reasons, 
we omit the definitions of the methods: 
\bruno{No need to show full code in the paper. You can just put \ldots in the body of the collectLit methods.}

@@

Note that the code for instantiation is automatically generated by \mixin. 
So, for example creating a simple expression of type \texttt{ExpPC} is 
as simple as:

\bruno{Add code here}

\noindent In contrast, in a pure Java solution, the tedious instantiation code would need 
to be defined manually.

\subsection{Embedded DSLs with Fluent Interfaces}
Since the style of fluent interfaces was invented in Smalltalk as method
cascading, more and more languages came to support fluent interfaces, including
JavaScript, Java, C++, D, Ruby, Scala, etc. For most languages, to create fluent
interfaces, programmers have to either handwrite everything or create a
wrapper around the original non-fluent interfaces using \textbf{this}. In Java,
several libraries help create fluent APIs easily (by code generation)\bruno{do they use code 
generation to produce the fluent interfaces?}, including
jOOQ, op4j, fluflu, JaQue, etc.

The \mixin annotation can also be used to create fluent interfaces. 
When creating fluent interfaces with \mixin, there are two main advantages:
\begin{enumerate}
\item Instead of forcing programmers to hand write code using \textbf{return
    this}, our approach with \mixin annotation removes this verbosity and
  automatically generate fluent setters.
\item Along the extension direction, the return types of fluent setters are
  automatically refined.
\end{enumerate}

\noindent We use embedded DSLs of two simple SQL query languages to illustrate.
The first query language \texttt{Database}  models
\texttt{select}, \texttt{from} and \texttt{where} clauses:
\lstinputlisting[linerange=6-13]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_DATABASE
\bruno{Wouldn't it be better to have a static method create() that calls of("","","")) and initializes all the fields? It seems awkward 
to have to use the of explicitly to initialized all fields.}

\noindent The main benefit that fluent methods give
us is the convinience of method chaining:

\lstinputlisting[linerange=35-38]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_QUERY1
\bruno{write query in 1 line}

\noindent Note how all the logic for the fluent setters is automatically provided by the \mixin annotation.

\paragraph{Extending the Query Language} The previous query language can be extended with a new feature
\texttt{orderBy} which orders the result records by a field that users
specify. With \mixin programmers just need to extend the interface \texttt{Database} with new
feature, and the return type of fluent setters in
\texttt{Database} is automatically refined to \texttt{ExtendedDatabase}:
\lstinputlisting[linerange=17-20]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_DATABASE_EXT
This way, when a \texttt{ExtendedDatabase} query created, 
all the fluent setters return the correct type, and not the old \texttt{Database} type, which would prevent calling 
\texttt{orderBy}.

\lstinputlisting[linerange=44-48]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_QUERY2
\bruno{write query in 1 line}

\subsection{A Small Game Example}
The last case study is a simple Maze game, which is used\bruno{where?}
to evaluate code reuse ability related to inheritance and design
patterns. In the game, there is a player with the goal of collecting
as many coins as possible. She may enter a room with several doors to
be chosen among. This is a good example because it involves code reuse
(different kinds of doors inherit a common type, with different
features and behaviour), multiple inheritance (a special kind of door
may require features from two other door types) and it also shows how
to model operations \texttt{symmetric sum}, \texttt{override} and
\texttt{alias} as trait-oriented programming. The game has been
implemented using plain Java 8 and default methods by Bono
et. al~\cite{bono14}, and the code for that implementation is
available online. We reimplemented the game using \mixin. Due to space
constraints, we omit the code here. The following table summarizes
the number of lines of code and classes/interfaces in each implementation:

\vspace{5pt}
\hspace{80pt}\begin{tabular}{ccc}
\hline
            & SLOC   & \# of classes/interfaces \\ \hline
Bono et al. & 335    & 14                       \\
Ours        & 199    & 11                       \\
\rowcolor[HTML]{C0C0C0} 
Reduced by  & 40.6\% & 21.4\%                   \\ \hline
\end{tabular}
\vspace{5pt}

\noindent The \mixin annotation allowed us to reduce the interfaces/classes used
in Bono et al.'s implementation by 21.4\% (from 14 to 11). The
reductions was due to the replacement of instantiation classes with
generated \texttt{of} methods. The number of source lines of code (SLOC)
was reduced by 40\%\bruno{explain why? we generate code for getters/setters?}. 
To ensure a fair comparison, we used the same coding style as Bono's.
